---
title: "discussion"
format: html
editor_options: 
  chunk_output_type: console
---

```{r}
library(tidyverse)
library(sf)
library(terra)
library(spData)
library(spDataLarge)
```

```{r}
# Load raster data representing grain sizes with the three classes clay, silt and sand
grain <- terra::rast(system.file("raster/grain.tif", package = "spData"))
```

Subset points higher than 3100 meters in nz_height (from spData) and create a template raster with rast(), where the resolution is 3 km x 3 km for the extent of the subset dataset

1. Count numbers of the highest points in each grid cell
2. Find the maximum elevation in each grid cell

```{r}
nz_height <- spData::nz_height
```


```{r}
# Subset points higher than 3100 meters
height_subset <- nz_height |>
  subset(elevation > 3100)

# Create a template raster with resolution is 3km x 3km
temp_rast <- rast(ext(height_subset), resolution = 3000, crs = crs(height_subset))
```

1. Count numbers of the highest points in each grid cell
```{r}
# Convert vector points to raster data
# Function "length" returns a count of the elevation points per cell
nz_raster <- rasterize(height_subset, temp_rast, field = "elevation", fun = "length")

plot(nz_raster, main = "Number of Elevation Points > 3100 in Each Grid Cell")
plot(st_geometry(height_subset), add = TRUE)
```

2. Find the maximum elevation in each grid cell
```{r}
# function "max" returns maximum elevation value per cell
nz_raster2 <- rasterize(height_subset, temp_rast, field = "elevation", fun = max)

plot(nz_raster2, main = "Maximum Elevation in Each Grid Cell ")
plot(st_geometry(height_subset), add = TRUE)
```


Aggregate the raster that counts the highest points in New Zealand/Aotearoa
```{r}
# Reduce the resolution by combining 2 cells in each direction into larger cells
# Sum the values of all cells for the resulting elevation value
nz_raster_low <- aggregate(nz_raster, fact = 2, fun = sum, na.rm = TRUE)

# Convert the new raster's resolution back to the 3kmx3km resolution of original raster
nz_resample <- resample(nz_raster_low, nz_raster)

plots <- c(nz_raster, nz_resample)
labs <- c("Original 6 x 6 km", "Resample 6 x 6 km")
plot(plots, main = labs)

```









